Efficient evaluation of threshold queries of derived fields in a numerical simulation database

### 2015-01-11 paper.tex
Added NSF grants under acknowledgments.
Minor changes and typo fixes.

\begin{abstract}
%Data-intensive architectures and compute clusters have been deployed to store, manage and provide public access to such datasets. 
%The systems built on top of these architectures, sometimes called virtual laboratories, usually do not support the analysis functionality that scientists are 
%interested in out of the box. Moreover, d
In this paper we present a method for the efficient evaluation of threshold queries of derived fields for large numerical 
simulation datasets stored in a cluster of relational databases. 
Such simulations are commonplace in the study of many different physical phenomena. 
The datasets produced by these simulations are in the TB and even PB ranges. 
Data-intensive computations that examine entire time-steps of the simulation data are impractical to perform
locally by the user taking days or months to iterate over the entire dataset. 
The integrated method for the evaluation of threshold queries that we have developed achieves scalability through data-parallel execution
of the computations on the nodes of an analysis database cluster. 
We extend the scientific analysis environment with the introduction of an application-aware cache for query results, building on the concept of semantic caching.
The cache has little overhead and improves query performance by over an order of magnitude for queries that hit the cache.
In the case of computational turbulence this allows scientists to quickly focus on the most intense events and interesting regions in any time-step or 
the dataset as a whole, which greatly speeds up the rate of scientific exploration and discovery. 
Furthermore, caching the results of threshold queries preserves both the I/O and computation effort used to obtain them, which allows subsequent
queries for some of the most interesting regions and features of the dataset to run orders of magnitude faster.
\end{abstract}

%% % A category with the (minimum) three required fields
\category{H.2.8}{Database Management}{Database Applications -- scientific, spatial databases and GIS}
%% %A category including the fourth, optional field follows...
%% \category{D.2.8}{Software Engineering}{Metrics}[complexity measures, performance measures]

%\terms{Extending Database Technology}

@@

\section{Introduction}
%What is the problem?
Better instruments, faster and bigger supercomputers and easier collaboration and sharing of data in the sciences have introduced the need
to manage increasingly large datasets. Data-intensive systems and architectures have been developed with the goal of storing and providing
fast access to such datasets. Examples of such analysis environments include the GrayWulf and Data-Scope clusters \cite{Szalay, DataScope} at 
Johns Hopkins, which have capacity of 1.1PB and 11PB
respectively. One of their missions is to provide persistent storage and public access to world-class numerical simulation data. 
These systems differ from the traditional HPC environments in that they aim to achieve high aggregate throughput by balancing computation
capabilities with I/O and network bandwidth. The computing systems and services developed on top of these platforms are more than pure
storage engines and usually have complex analysis routines built-in, which has largely been driven by the ``move the computation to the data"
paradigm \cite{fourthparadigm}. These built-in analysis routines are most often not novel themselves. They implement core scientific functionality for the
study of the particular scientific phenomena, which was observed or simulated in the first place. The analysis routines however require novel evaluation
strategies and methods for their execution. They have to operate on large array datasets distributed across multiple nodes of a cluster of relational
databases. In order to reduce their running times they have to both make efficient use of the cluster resources and incorporate 
leading data management techniques.

Finding the locations or regions of highest vorticity or those with the largest norms of the velocity or other fields of interest is
functionality that enables new insights in the study of fluid dynamics. Analysis of this kind coupled with the ability to analyze time-series datasets both 
forward and backward in time has transformed our understanding of turbulence \cite{Eyink}. Furthermore, threshold, top-$k$ queries and similarity search
in general are important in many different disciplines. We introduce an efficient evaluation strategy for threshold
queries over time-series datasets stored in a cluster of relational databases. Our method evaluates not only threshold queries of the vector or 
scalar field data stored in the database, but also performs thresholding of \emph{derived} fields.

The main challenge that our approach tackles is that the
data-intensive computation of \emph{derived} fields has to be carried out on-demand for extremely large array datasets stored in an analysis cluster environment 
comprised of multiple database nodes. We focus on the evaluation of threshold queries of fields derived from the data stored in the cluster as
these queries are the most interesting scientifically. However, our approach applies to the evaluation of top-$k$ queries, rollup queries and data-reducing 
queries in general. The established data management techniques that our approach combines make it easy to understand. It can be applied to other 
scientific analysis environments, which manage large datasets in a database management system. Examples include the Sloan Digital Sky Survey
\cite{SDSS}, the Millennium Simulation \cite{Millennium} and the Open Connectome Project \cite{Burns}.

@@

Database, operating and file system caches are effective at speeding up access to the large amounts of data stored on disk. However, this might not be
sufficient for some interactive applications, because these application-independent caches cannot exploit dataset-specific structure and application-level
information \cite{Lopez}. Moreover, even if the data are available in one or more of these application-independent caches the computation associated 
with the derived field still needs to be performed for each point on the grid, because results of previous computations are not cached. 
We will demonstrate that an integrated approach, which computes the derived fields on-demand in a data-parallel manner, performs the evaluation over 
an entire time-step in a few minutes. Storing the query results in an application-aware semantic cache further reduces the running times to several 
seconds.

@@

%Why hasn't it been solved before? (Or, what's wrong with previous proposed solutions? How does mine differ?)
There are several challenges that arise during the evaluation of threshold queries of derived fields in an analysis database cluster. The field variables have to be
evaluated \emph{on-demand} from the array data stored in the database cluster. The computations are data-intensive as they apply stencil kernels on extremely
large multidimensional array datasets. Stencil kernel computations have to be performed at each location on the grid as opposed to at a particular number 
of target locations. 
The evaluation needs to be distributed across the nodes of the database cluster to avoid the unnecessary movement of data over 
the network and to achieve scalability.
Techniques that target the traditional supercomputing environments do not translate directly to the 
distributed database setting of an analysis cluster environment.

@@

%What are the key components of my approach and results? Also include any specific limitations.
We present a method for the efficient evaluation of threshold queries over fields derived from the raw vector or scalar fields of the numerical simulation 
stored in the database. The derived fields, which our method computes and operates on are derived by performing stencil kernel computations over the 
stored array data.
Our method makes effective use of the cluster resources and achieves high throughput and scalability. 
We exploit the parallelism 
available in the cluster by means of data-parallel execution of the computations. 
%Each computation is distributed across the nodes of the cluster and
%is executed in parallel by several processes within a database node by creating multiple connections to each node. 
We extend the database management system with an
application-aware cache for query results. We build on the idea of an application-aware cache introduced by Lopez et al. \cite{Lopez} and more broadly 
on the concept of semantic caching \cite{Dar}. Rather than caching just data as is the case in system caches and the tree-cache described by Lopez et al.,
we cache the actual results of the computation in a limited version of a semantic cache. 
This leads to query performance improvement of over an order of magnitude.

The contributions of this paper are the following:
\begin{itemize}
\item Computing derived fields of large simulation data on-demand and evaluating threshold queries on them at extreme scale. For the first time providing large
data analytics capabilities that examine entire time-steps of the simulation transparently to the user in a production analysis environment.
\item Achieving this through the combination of existing data management techniques such as data parallelism and semantic caching as well as taking
advantage of heterogeneous scientific cluster architectures (sharded relational DBMS with several SSDs per node).
\item Evaluating the proposed method on data-intensive workloads in a live production environment and showing scalability results on datasets hundreds of terabytes in size.
\end{itemize}

@@

\section{JHTDB}\label{jhtdb}
Data-intensive architectures and compute clusters built from commodity hardware rely on parallel I/O to multiple disks and high network bandwidth
to achieve high throughput. Such systems have only recently been deployed for the storage of large numerical simulation 
datasets. The virtual laboratories built on these systems make use of relational database system technology to store and manage these large array datasets.
Relational database systems however often do not support
all of the functionality that scientists are interested in out of the box. It is either up to the user to develop more sophisticated analysis routines locally or such
capabilities have to be built-in through user-defined functions or stored procedures.

@@

The database nodes are part of the GrayWulf \cite{Szalay} and Data-Scope \cite{DataScope} clusters. Each node
is running Windows Server 2008 and SQL Server 2008 R2. The data for each dataset reside on a regular three-dimensional spatial grid with the
exception of the channel flow data, which has an irregular $y$ dimension. The data are partitioned spatially across 4 to 8
database nodes, and each database node hosts one or more databases. We use the Morton z-order space-filling
curve to distribute the data across nodes and databases. Each time-step is spatially subdivided
into database atoms, which are of size $8^3$. Each such atom is indexed by the time-step, which it belongs to and
by the Morton code of it's lower left corner. This combination of index and data forms a record in the database.
Queries to the data and derived fields, such as derivatives and filtered quantities are evaluated through
stored procedures or user-defined functions implemented in the Common Language Runtime (CLR) framework.

@@

The JHTDB provides a variety of data-intensive analysis routines that are executed on the database nodes. These include interpolation, 
differentiation, particle tracking and spatial filtering. These tasks are often data-intensive and in order to leverage the capabilities of the cluster we
have developed data-driven batch processing techniques for their evaluation \cite{KanovSC11, KanovSC12}. Most of these tasks usually operate on subsets
of the space or a collection of individual target locations within a time-step. In contrast, threshold and top-$k$ queries usually have to examine the entire 
data volume of a time-step or a signification portion of it. Furthermore, the data product of threshold queries is much smaller in relation to the amount of 
data that need to be examined. This fact combined with the fact that subsequent queries can reuse 
previously computed results makes the query results suitable to caching.

@@

\begin{figure*}[t]
\centering
\includegraphics[width=7in]{Figures/most_intense_cluster_alt.pdf}
\caption{3D (fixed frame) cut through the 4D cluster containing the most intense event.}
\label{fig:max_vort_cluster}
\end{figure*}

@@

In addition to obtaining the regions of largest vorticity, there is substantial interest in studying the regions with highest values for other fields, such as
the second and third velocity gradient invariants ($Q$ and $R$). These invariants are scalar quantities whose values contain information about the 
topology of the flow and the rates of vortex stretching and rotation. In MHD, finding the locations with largest values for the electric current
can lead to new insights into the development of the most intense reconnection events of magnetic field sheets in the simulated plasma. Similarly to
the vorticity, the electric current is derived from the magnetic field by taking its curl. The list of fields of interest, which scientists would like to perform 
threshold queries on certainly does not stop here and is indicative of how valuable this functionality is in the study of turbulence and fluid dynamics.

@@

{\bf Distributed data-parallel execution:} In most cases threshold queries operate over an entire time-step. Each such query is subdivided by 
the mediator into queries submitted to
each of the database nodes. Each node evaluates the query over the data that it has stored locally. Only a small amount of data along the boundary
need to be requested from adjacent nodes. The size of the band of data that may not be available locally is equal to a kernel half-width. 
Such a band is needed on each of the sides of the box forming the domain of the computation.  
The data are read into memory and the particular field requested is computed at each of the locations on
the grid. The same strategy applies when utilizing multiple processes per node. The norm or absolute value of the derived field at each location is compared 
against the specified threshold and if the value is higher it is maintained along with
the spatial coordinates of the location in a list.

@@

The cached query results are stored in a table in the database and the overall size of the cache is limited by the amount of available disk space, not memory.
Given a limit of $\sim$$10^6$ points per time-step for a threshold query, the space required to cache the maximum number of points including the index
space and database overhead is $\sim$40MB. 
Therefore for a dataset containing 
1024 time-steps, as is the case for the isotropic turbulence and MHD datasets part of the JHTDB, a cache size of 40GB is sufficient to cache the query
results for threshold queries of a derived field over the entire dataset. In contrast, computing and materializing a scalar derived field for the entire dataset will
require $\sim$5TB (15TB for vector fields).

@@

\begin{figure}
\centering
\includegraphics[width=3.35in]{Figures/cache_overhead.pdf}
\caption{Semi-log plot of the execution time for threshold queries at different threshold levels compared with the execution time of the same queries in the
absence of a cache.}
\label{fig:cache_overhead}
\end{figure}

@@

\subsection{Scaling and Distributed Evaluation}
The evaluation of threshold queries of derived fields from the raw data is both I/O and computationally bound. These queries examine the entire data
volume of a simulation time-step and are therefore good candidates for a data-parallel distributed evaluation. 
Our data-parallel implementation exhibits good vertical and nearly ideal horizontal scaling as shown in Figure \ref{fig:scaling}. 
For the scale-up (Fig. \ref{fig:scaleup}) experiments we used the same
queries and threshold values as for the runs shown in Figure \ref{fig:cache_overhead} and Table \ref{tab:cache_effectiveness} 
but with varying number of processes per node. Cache entries for the
time-step queried were again dropped before each run in order to evaluate the scaling properties of the computation of the derived field from the stored data.
The computations for all of 
the derived fields of interest (such as the vorticity) at each grid point need data from adjacent grid points only. Therefore, each node of the cluster is able to
compute the derived field from data available locally with only a small amount along the boundary of each region having to be retrieved from adjacent nodes.
Each computation is independent and embarrassingly parallel. This allows us to make use of multiple processes per node and scale out to
multiple database nodes.

@@

In all of these cases the time taken to interrogate the cache is negligible. The mediator time to distribute the queries and assemble the results as well as the
time to transfer them to the user are also substantially smaller than the I/O and computation times. As expected they increase proportionally to the number
of points in the result set.

It is interesting to note that the time taken to perform a cache lookup is relatively small even in the case of a cache hit as can be seen in Figures 
\ref{fig:vorticity_breakdown_cachehit}, \ref{fig:q_breakdown_cachehit}, \ref{fig:magnetic_breakdown_cachehit}. This is because the cache tables reside on
SSDs attached to each database node (see Fig. \ref{fig:cache_diagram}) and retrieving the data is always done through a clustered index lookup.
In the cases of a cache hit the majority of the time is spent simply transferring the results from the database nodes to the mediator and then back to the user.
These times remain more or less constant between the cases of cache misses and cache hits (top row and bottom row of Fig. \ref{fig:breakdown}).
Caching the results of threshold queries
effectively preserves the I/O and computational effort spent during their initial evaluation and results in over an order of magnitude speedup for all the
different fields requested as we can see in Figure \ref{fig:breakdown}.

\vspace{2mm}
\section{Related Work}
Only select few database systems offer support for arrays as first-class citizens. Even fewer provide the fault-tolerance, scalability and availability guarantees 
necessary for a system managing multi-terabyte datasets in a production setting. This is part of the reason why we have chosen to represent the array
data in the JHTDB as a collection of binary large objects in a relational DBMS and perform the array manipulation tasks necessary at the application level.
The systems that provide
support for arrays and aim to handle large array data efficiently are RasDaMan \cite{Rasdaman}, SciDB \cite{Scidb} and MonetDB/SciQL \cite{Sciql}. RasDaMan
partitions raster objects into tiles, which are stored in a traditional relational database system. This approach is similar to how the numerical simulation data
are handled in the JHTDB. RasDaMan provides RasQL \cite{Rasql}, which is a SQL-92 based query language for the manipulation of raster images. SciDB
is an array database system build from the ground up. Array attributes are partitioned vertically and each attribute array is decomposed into overlapping
chunks. SciDB provides a declarative Array Query Language (AQL) and an Array Functional Language (AFL). Users can create arrays with named
dimensions with AQL and make use of the functional operators defined in AFL, such
as SLICE, SUBSAMPLE, SJOIN, FILTER and APPLY. SciQL's focus is on language design and integration with SQL:2003 syntax and semantics. It is implemented
within the MonetDB framework \cite{MonetDB}.

Database systems support rollup queries, including top-$k$ queries, but in most cases these
queries apply only simple linear score functions on the attribute values of individual records.
Additionally, many top-$k$ query evaluation techniques rely on the score functions being monotone in order to perform early pruning (see \cite{Ilyas} for a
survey of top-$k$ evaluation strategies).
This is an assumption that we cannot make for the functions used to compute all the different possible derived fields of interest in fluid mechanics.
Even approaches that aim to work with general score functions \cite{Deshpande, Xin} assume that the function operates on the attributes of a single record.
In contrast our approach performs a stencil kernel computation at each grid location in order to obtain the value of a \emph{derived} field at that location and
examines the vector or scalar array data at all neighboring locations, which
are within the kernel of the computation. The functions used to derive the field may even be non-linear.
Finally, a top-$k$ approach may not be suitable in the cases where scientists are interested in performing threshold queries at different time-steps as the
same threshold level will produce different number of points in the result set for different time-steps.

@@

In a system called the tree cache Lopez et al. \cite{Lopez} make use of a small application-aware cache to reduce access time to large datasets stored on
disk. The tree cache stores individual octants of octree datasets and exploits application-specific information to determine which octants to cache and to 
perform query reordering. This work has inspired the use of an application-aware cache for the evaluation of threshold queries. In contrast to the tree cache,
we do not cache raw data objects, but rather query results. Caching query results preserves the computational effort in addition to reducing I/O, which has a
much bigger impact on query performance and substantially reduces the size of the cache. Additionally, the cache that we introduce resides on disk rather 
than in memory, which greatly increases its potential size. Lopez et al. also explore approximate querying through aggregation, which can be fairly easily
supported by our system but is of limited use as scientists performing threshold queries are usually interested in obtaining the exact locations where a field is 
at its highest values.

@@

\section{Acknowledgments}
The authors would like to thank the Turbulence Database Group at Johns Hopkins University as well as three anonymous reviewers for
their insightful comments and suggestions. This work is supported in part by the National Science Foundation
under Grants to..

>>>>

\begin{abstract}
%Data-intensive architectures and compute clusters have been deployed to store, manage and provide public access to such datasets. 
%The systems built on top of these architectures, sometimes called virtual laboratories, usually do not support the analysis functionality that scientists are 
%interested in out of the box. Moreover, d
In this paper, we present a method for the efficient evaluation of threshold queries of derived fields for large numerical 
simulation datasets stored in a cluster of relational databases. 
%Such simulations are commonplace in the study of many different physical phenomena. 
The datasets produced by these simulations are in the TB and even PB ranges. 
Data-intensive computations that examine entire time-steps of the simulation data are impractical to perform
locally by the user, taking days or months to iterate over the entire dataset. 
The integrated method for the evaluation of threshold queries that we have developed achieves scalability through data-parallel execution
of the computations on the nodes of an analysis database cluster. 
We extend the scientific analysis environment with the introduction of an application-aware cache for query results, building on the concept of semantic caching.
The cache has little overhead and improves query performance by over an order of magnitude for queries that hit the cache.
Caching the results of threshold queries preserves both the I/O and computation effort used to obtain them.
In the case of computational turbulence, this allows scientists to quickly focus on the most intense events and interesting regions in any time-step or 
the dataset as a whole, which greatly speeds up the rate of scientific exploration and discovery. 
\end{abstract}

%% % A category with the (minimum) three required fields
\category{H.2.8}{Database Management}{Database Applications -- Scientific Databases}
\category{H.2.4}{Database Management}{Systems -- Distributed Databases}
\category{J.2}{Computer Applications}{Physical Sciences and Engineering -- Physics}

@@

\vspace{10pt}
\section{Introduction}
%What is the problem?
Better instruments, faster and bigger supercomputers and easier collaboration and sharing of data in the sciences have introduced the need
to manage increasingly large datasets. Data-intensive systems and architectures have been developed with the goal of storing and providing
fast access to such datasets. Examples of such analysis environments include the GrayWulf and Data-Scope clusters \cite{Szalay, DataScope} at 
Johns Hopkins, which have capacity of 1.1PB and 11PB
respectively. One of their missions is to provide persistent storage and public access to world-class numerical simulation data. 
These systems differ from the traditional HPC environments in that they aim to achieve high aggregate throughput by balancing computation
capabilities with I/O and network bandwidth. The computing systems and services developed on top of these platforms are more than pure
storage engines and usually have complex analysis routines built-in, which has largely been driven by the ``move the computation to the data"
paradigm \cite{fourthparadigm}. These built-in analysis routines are most often not novel themselves. They implement core scientific functionality for the
study of the particular scientific phenomena, which was observed or simulated in the first place. The analysis routines however require novel evaluation
strategies and methods for their execution. They have to operate on large array datasets distributed across multiple nodes of a cluster of relational
databases. In order to reduce their running times, they have to make efficient use of the cluster resources and incorporate 
leading data management techniques.

Finding the locations or regions of highest vorticity or those with the largest norms of the velocity or other fields of interest
enables new insights in the study of fluid dynamics. Analysis of this kind coupled with the ability to analyze time-series datasets both 
forward and backward in time has transformed our understanding of turbulence \cite{Eyink}. Furthermore, threshold, top-$k$ queries and similarity search
in general are important in many different disciplines. We introduce an efficient evaluation strategy for threshold
queries over time-series datasets stored in a cluster of relational databases. Our method evaluates not only threshold queries of the vector or 
scalar field data stored in the database, but also performs thresholding of \emph{derived} fields.

The main challenge that our approach tackles is that the
data-intensive computation of \emph{derived} fields has to be carried out on-demand for extremely large array datasets stored in an analysis cluster environment 
comprised of multiple database nodes. We focus on the evaluation of threshold queries of fields derived from the data stored in the cluster as
these queries are the most interesting scientifically. However, our approach applies to the evaluation of top-$k$ queries, rollup queries and data-reducing 
queries in general. The established data management techniques that our approach combines make the approach easy to understand. It can be applied to other 
scientific analysis environments, which manage large datasets in a database management system. Examples include the Sloan Digital Sky Survey
\cite{SDSS}, the Millennium Simulation \cite{Millennium} and the Open Connectome Project \cite{Burns}.

@@

Database, operating and file system caches are effective at speeding up access to the large amounts of data stored on disk. However, this might not be
sufficient for some interactive applications, because these application-independent caches cannot exploit dataset-specific structure and applica\-tion-level
information \cite{Lopez}. Moreover, even if the data are available in one or more of these application-independent caches the computation associated 
with the derived field still needs to be performed for each point on the grid, because results of previous computations are not cached. 
We will demonstrate that an integrated approach, which computes the derived fields on-demand in a data-parallel manner, performs the evaluation over 
an entire time-step in a few minutes. Storing the query results in an application-aware semantic cache further reduces the running times to several 
seconds.

@@

%Why hasn't it been solved before? (Or, what's wrong with previous proposed solutions? How does mine differ?)
There are several challenges that arise during the evaluation of threshold queries of derived fields in an analysis database cluster. The field variables have to be
evaluated \emph{on-demand} from the array data stored in the database cluster. The computations are data-intensive as they apply stencil kernels on extremely
large multidimensional array datasets. 
Stencil kernel computations have to be performed at each location on the grid as opposed to at a particular number of target locations. 
Each data point appears in $N$ kernels of computation, in which $N$ is the size of the kernel.
The evaluation needs to be distributed across the nodes of the database cluster to avoid the unnecessary movement of data over 
the network and to achieve scalability.
Techniques that target the traditional supercomputing environments do not translate directly to the 
distributed database setting of an analysis cluster environment.

@@

%What are the key components of my approach and results? Also include any specific limitations.
We present a method for the efficient evaluation of threshold queries over fields derived from the raw vector or scalar fields of the numerical simulation 
stored in the database. 
Our method makes effective use of the cluster resources and achieves high throughput and scalability. 
We exploit the parallelism 
available in the cluster by means of data-parallel execution of the computations. 
%Each computation is distributed across the nodes of the cluster and
%is executed in parallel by several processes within a database node by creating multiple connections to each node. 
We extend the database management system with an
application-aware cache for query results. We build on the idea of an application-aware cache introduced by Lopez et al. \cite{Lopez} and more broadly 
on the concept of semantic caching \cite{Dar}. Rather than caching just data as is the case in system caches and the tree-cache described by Lopez et al.,
we cache the actual results of the computation in a limited version of a semantic cache. 
This leads to query performance improvement of over an order of magnitude.

The contributions of this paper are the following:
\begin{itemize}
\item Computing derived fields of large simulation data on-demand and evaluating threshold queries on them at extreme scale. This provides large
data analytics capabilities that examine entire time-steps of the simulation transparently to the user in a production analysis environment.
\item Achieving this through the combination of existing data management techniques such as data parallelism and semantic caching as well as taking
advantage of heterogeneous scientific cluster architectures (sharded relational DBMS with several SSDs per node).
\item Evaluating the proposed method on data-intensive workloads in a live production environment and showing scalability results on datasets hundreds of terabytes in size.
\end{itemize}

@@

\section{JHTDB}\label{jhtdb}
Data-intensive architectures and compute clusters built from commodity hardware rely on parallel I/O to multiple disks and high network bandwidth
to achieve high throughput. Such systems have only recently been deployed for the storage of large numerical simulation 
datasets. The virtual laboratories built on these systems make use of relational database system technology to store and manage large array datasets.
Relational database systems however often do not support
all of the functionality that scientists are interested in out of the box. It is either up to the user to develop more sophisticated analysis routines locally or such
capabilities have to be built into the database through user-defined functions or stored procedures.

@@

The database nodes are part of the GrayWulf \cite{Szalay} and Data-Scope \cite{DataScope} clusters. Each node
is running Windows Server 2008 and SQL Server 2008 R2. The data for each dataset reside on a regular three-dimensional spatial grid with the
exception of the channel flow data, which has an irregular $y$ dimension. The data are partitioned spatially across 4 to 8
database nodes, and each database node hosts one or more databases. We use the Morton z-order space-filling
curve to distribute the data across nodes and databases \cite{Perlman}. Each time-step is spatially subdivided
into database atoms, which are of size $8^3$. Each such atom is indexed by the time-step, which it belongs to and
by the Morton code of it's lower left corner. This combination of index and data forms a record in the database.
Queries to the data and derived fields, such as derivatives and filtered quantities are evaluated through
stored procedures or user-defined functions implemented in the Common Language Runtime (CLR) framework.

@@

The JHTDB provides a variety of data-intensive analysis routines that are executed on the database nodes. These include interpolation, 
differentiation, particle tracking and spatial filtering. These tasks are often data-intensive and in order to leverage the capabilities of the cluster we
have developed data-driven batch processing techniques for their evaluation \cite{KanovSC11, KanovSC12}. Most of these tasks usually operate on subsets
of the space or a collection of individual target locations within a time-step.

In contrast, threshold and top-$k$ queries usually have to examine the entire 
data volume of a time-step or a signification portion of it. Furthermore, the data product of threshold queries is much smaller in relation to the amount of 
data that need to be examined. This fact combined with the fact that subsequent queries can reuse 
previously computed results makes the query results suitable to caching.

@@

\begin{figure*}[t]
\centering
\includegraphics[width=7in]{Figures/most_intense_cluster_alt.pdf}
\caption{3D (single time-step) cut through the 4D cluster containing the most intense event.}
\label{fig:max_vort_cluster}
\end{figure*}

@@

In addition to obtaining the regions of largest vorticity, there is substantial interest in studying the regions with highest values for other fields, such as
the second and third velocity gradient invariants ($Q$ and $R$). These invariants are scalar quantities whose values contain information about the 
topology of the flow and the rates of vortex stretching and rotation. In MHD, finding the locations with largest values for the electric current
can lead to new insights into the development of the most intense reconnection events of magnetic field sheets in the simulated plasma. Similarly to
the vorticity, the electric current is derived from the magnetic field by taking its curl. The list of fields of interest, on which scientists would like to perform 
threshold queries certainly does not stop here and is indicative of how valuable this functionality is in the study of turbulence and fluid dynamics.

@@

{\bf Distributed data-parallel execution:} In most cases threshold queries operate over an entire time-step. Each such query is subdivided by 
the mediator into queries submitted to
each of the database nodes. Each node evaluates the query over the data that it has stored locally. Only a small amount of data along the boundary
need to be requested from adjacent nodes. The size of the band of data that may not be available locally is equal to a kernel half-width. 
Such a band is needed on each of the sides of the box forming the domain of the computation.  
The data are read into memory and the particular field requested is computed at each of the locations on
the grid. The same strategy applies when utilizing multiple processes per node. The norm or absolute value of the derived field at each location is compared 
against the specified threshold and if the value is higher, it is maintained along with
the spatial coordinates of the location in a list.

@@

The cached query results are stored in a table in the database and the overall size of the cache is limited by the amount of available SSD disk space, not memory.
Given a limit of $\sim$$10^6$ points per time-step for a threshold query, the space required to cache the maximum number of points including the index
space and database overhead is $\sim$40MB. 
Therefore for a dataset containing 
1024 time-steps, as is the case for the isotropic turbulence and MHD datasets part of the JHTDB, a cache size of 40GB is sufficient to cache the query
results for threshold queries of a derived field over the entire dataset. In contrast, computing and materializing a scalar derived field for the entire dataset will
require $\sim$5TB (15TB for vector fields).

@@

\begin{figure}
\centering
\includegraphics[width=3.35in]{Figures/cache_overhead.pdf}
\caption{Execution time for threshold queries at different threshold levels compared with the execution time of the same queries in the
absence of a cache.}
\label{fig:cache_overhead}
\end{figure}

@@

\subsection{Scaling and Distributed Evaluation}
The evaluation of threshold queries of derived fields from the raw data is both I/O and computationally bound. These queries examine the entire data
volume of a simulation time-step and are, therefore, good candidates for a data-parallel distributed evaluation. 
Our data-parallel implementation exhibits good vertical and nearly ideal horizontal scaling as shown in Figure \ref{fig:scaling}. 
For the scale-up experiments (Fig. \ref{fig:scaleup}), we used the same
queries and threshold values as for the runs shown in Figure \ref{fig:cache_overhead} and Table \ref{tab:cache_effectiveness} 
but with varying number of processes per node. Cache entries for the
time-step queried were again dropped before each run in order to evaluate the scaling properties of the computation of the derived field from the stored data.
The computations for all of 
the derived fields of interest (such as the vorticity) at each grid point need data from adjacent grid points only. Therefore, each node of the cluster is able to
compute the derived field from data available locally with only a small amount along the boundary of each region having to be retrieved from adjacent nodes.
Each computation is independent and embarrassingly parallel. This allows us to make use of multiple processes per node and scale out to
multiple database nodes.

@@

In all of these cases, the time taken to interrogate the cache is negligible. The mediator time to distribute the queries and assemble the results as well as the
time to transfer them to the user are also substantially smaller than the I/O and computation times. As expected they increase proportionally to the number
of points in the result set.

It is interesting to note that the time taken to perform a cache lookup is relatively small even in the case of a cache hit as can be seen in Figures 
\ref{fig:vorticity_breakdown_cachehit}, \ref{fig:q_breakdown_cachehit} and \ref{fig:magnetic_breakdown_cachehit}. This is because the cache tables reside on
SSDs attached to each database node (see Fig. \ref{fig:cache_diagram}) and retrieving the data is always done through a clustered index lookup.
In the cases of a cache hit, the majority of the time is spent simply transferring the results from the database nodes to the mediator and then back to the user.
These times remain more or less constant between the cases of cache misses and cache hits (top row and bottom row of Fig. \ref{fig:breakdown}).
Caching the results of threshold queries
effectively preserves the I/O and computational effort spent during their initial evaluation and results in over an order of magnitude speedup for all the
different fields requested as we can see in Figure \ref{fig:breakdown}.

\section{Related Work}
Only select few database systems offer support for arrays as first-class citizens. Even fewer provide the fault-tolerance, scalability and availability guarantees 
necessary for a system managing multi-terabyte datasets in a production setting. This is part of the reason why we have chosen to represent the array
data in the JHTDB as a collection of binary large objects in a relational DBMS and perform the array manipulation tasks necessary at the application level.
The systems that provide
support for arrays and aim to handle large array data efficiently are RasDaMan \cite{Rasdaman}, SciDB \cite{Scidb} and MonetDB/SciQL \cite{Sciql}. RasDaMan
partitions raster objects into tiles, which are stored in a traditional relational database system. This approach is similar to how the numerical simulation data
are handled in the JHTDB. RasDaMan provides RasQL \cite{Rasql}, which is a SQL-92 based query language for the manipulation of raster images. SciDB
is an array database system build from the ground up. Array attributes are partitioned vertically and each attribute array is decomposed into overlapping
chunks. SciDB provides a declarative Array Query Language (AQL) and an Array Functional Language (AFL). Users can create arrays with named
dimensions with AQL and make use of the functional operators defined in AFL, such
as SLICE, SUBSAMPLE, SJOIN, FILTER and APPLY. SciQL's focus is on language design and integration with SQL:2003 syntax and semantics. It is implemented
within the MonetDB framework \cite{MonetDB}.

Database systems support rollup queries, including top-$k$ queries, but in most cases these
queries apply only simple linear score functions on the attribute values of individual records.
Additionally, many top-$k$ query evaluation techniques rely on the score functions being monotone in order to perform early pruning (see \cite{Ilyas} for a
survey of top-$k$ evaluation strategies).
This is an assumption that we cannot make for the functions used to compute all the different possible derived fields of interest in fluid mechanics.
Even approaches that aim to work with general score functions \cite{Deshpande, Xin} assume that the function operates on the attributes of a single record.
In contrast, our approach performs a stencil kernel computation at each grid location in order to obtain the value of a \emph{derived} field at that location and
examines the vector or scalar array data at all neighboring locations, which
are within the kernel of the computation. The functions used to derive the field may even be non-linear.
Finally, a top-$k$ approach may not be suitable in the cases where scientists are interested in performing threshold queries at different time-steps as the
same threshold level will produce different number of points in the result set for different time-steps.

@@

In a system called the tree cache, Lopez et al. \cite{Lopez} make use of a small application-aware cache to reduce access time to large datasets stored on
disk. The tree cache stores individual octants of octree datasets and exploits application-specific information to determine which octants to cache and to 
perform query reordering. This work has inspired the use of an application-aware cache for the evaluation of threshold queries. In contrast to the tree cache,
we do not cache raw data objects, but rather query results. Caching query results preserves the computational effort in addition to reducing I/O, which has a
much bigger impact on query performance and substantially reduces the size of the cache. Additionally, the cache that we introduce resides on disk rather 
than in memory, which greatly increases its potential size. Lopez et al. also explore approximate querying through aggregation, which can be fairly easily
supported by our system but is of limited use as scientists performing threshold queries are usually interested in obtaining the exact locations where a field is 
at its highest values.

@@

\section{Acknowledgments}
The authors would like to thank the Turbulence Database Group at Johns Hopkins University as well as three anonymous reviewers for
their insightful comments and suggestions. This work is supported in part by the National Science Foundation
under Grants CDI-II: CMMI 0941530 and OCI-108849 and Johns Hopkins University's Institute for Data Intensive Engineering \& Science.
